Analysis-ready VCF at Biobank scale using Zarr

### 2024-06-07 paper.tex
Eric's update
Tweaks to GEL section
Closes #140

To demonstrate the computational accessibility of Zarr on this 
large human dataset, we performed some illustrative benchmarks.
As these benchmarks take some time to run, we focus 
on a single 132GiB compressed VCF file covering
positions 58,219,159--60,650,943 (562,640 variants) 
from the middle of the set of 106 files for chromosome 2.
We report both the total CPU time and elapsed wall-clock time here
as both are relevant.
First, we extracted the genome position for each variant in this single VCF 
chunk using \texttt{bcftools query} and Python Zarr code as described in 
Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required 
% def parse_time(time_str):
%     minsplit = time_str.split("m")
%     return datetime.timedelta(minutes=int(minsplit[0]),
%         seconds=float(minsplit[1][:-1]))
% def parse_time_output(s):
%     values = {}
%     for line in s.splitlines():
%         split = line.split()
%         values[split[0]] = parse_time(split[1])
%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
% real  85m51.013s
% user  54m39.986s
% sys 0m45.277s
55.42 min CPU and 85.85 min elapsed.
% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
% Wall time: 1min 44s
% 55.42 * 60 / 2.78  = 1196
% 85.85 / 1.73 = 49.62 
The Zarr code required 2.78 sec CPU and 1.73 min elapsed.
This is a 1196X smaller CPU burden and a 50X speed-up in elapsed time.
The major difference between CPU time and wall-time is noteworthy 
here, and indicates some opportunities for improvement in VCF Zarr
in high-latency environments such as the shared file system in the
Genomics England HPC system. Currently VCF Zarr does not store any
specialised index to map genomic coordinates to array positions
along the variants dimension. Instead, to find the relevant slice
of records corresponding to the range of positions in the target
VCF file, we read in the entire variant\_position array and 
binary search. This entails reading 5,989 chunk files
(the chunk size is 100,000 variants) which incurs a substantial
latency penalty on this system. Later versions of the specification
may solve this problem by either storing a combined variant\_contig,
variant\_position array as a single chunk, or use a specialised
index that supports overlap queries.

@@

We then ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
and~\ref{fig-subset-matrix-compute}) on the VCF file
\texttt{bcftools +af-dist} as before. 
The elapsed time for this operation was 716.28 min CPU, 716.3 min elapsed.
Repeating this operation for the same coordinates in Zarr 
(using Python code described in previous sections) 
% 716.28 / 2.32 = 308.74
% 716.3 / 4.25 = 168.54
with a total CPU time of 2.32 min and elapsed time of 4.25 min.
This is a 309X reduction in CPU burden and a 169X speed-up in elapsed time.

@@

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} provide powerful abstractions
for large-scale data analysis and can be
used to perform interactive analysis
of large VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which optimised kernels are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis.

>>>>

To demonstrate the computational accessibility of Zarr on this 
large human dataset, we performed some illustrative benchmarks.
As these benchmarks take some time to run, we focus 
on a single 132GiB compressed VCF file covering
positions 58,219,159--60,650,943 (562,640 variants) 
from the middle of the list of 106 files for chromosome 2.
We report both the total CPU time and elapsed wall-clock time here
as both are relevant.
First, we extracted the genome position for each variant in this single VCF 
chunk using \texttt{bcftools query} and Python Zarr code as described in 
Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required 
% def parse_time(time_str):
%     minsplit = time_str.split("m")
%     return datetime.timedelta(minutes=int(minsplit[0]),
%         seconds=float(minsplit[1][:-1]))
% def parse_time_output(s):
%     values = {}
%     for line in s.splitlines():
%         split = line.split()
%         values[split[0]] = parse_time(split[1])
%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
% real  85m51.013s
% user  54m39.986s
% sys 0m45.277s
55.42 min CPU and 85.85 min elapsed.
% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
% Wall time: 1min 44s
% 55.42 * 60 / 2.78  = 1196
% 85.85 / 1.73 = 49.62 
The Zarr code required 2.78 sec CPU and 1.73 min elapsed.
This is a 1196X smaller CPU burden and a 50X speed-up in elapsed time.
The major difference between CPU time and wall-time is noteworthy 
here, and indicates some opportunities for improvement in VCF Zarr
in high-latency environments such as the shared file system in the
Genomics England HPC system. Currently VCF Zarr does not store any
specialised index to map genomic coordinates to array positions
along the variants dimension. Instead, to find the relevant slice
of records corresponding to the range of positions in the target
VCF file, we load the entire variant\_position array and 
binary search. This entails reading 5,989 chunk files
(the chunk size is 100,000 variants) which incurs a substantial
latency penalty on this system. Later versions of the specification
may solve this problem by storing an array of size 
(approximately) the number variant chunks 
which maps ranges of genome coordinates to chunk indexes,
or a more specialised structure that supports overlap queries.

@@

We then ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
and~\ref{fig-subset-matrix-compute}) on the VCF file
using \texttt{bcftools +af-dist} as before. 
The elapsed time for this operation was 716.28 min CPU, 716.3 min elapsed.
Repeating this operation for the same coordinates in Zarr 
(using Python code described in previous sections) 
% 716.28 / 2.32 = 308.74
% 716.3 / 4.25 = 168.54
gave a total CPU time of 2.32 min and elapsed time of 4.25 min.
This is a 309X reduction in CPU burden and a 169X speed-up in elapsed time.
It is worth noting here that \texttt{bcftools +af-dist} cannot be 
performed in parallel across multiple slices of a chromosome, and if 
we did want to run it on all of chromosome 2 we would need to 
concatenate the 106 VCF files. While af-dist itself is not a common operation,
many tasks share this property
of not being straightforwardly decomposable across multiple VCF files.

@@

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Dask~\cite{rocklin2015dask} and 
Xarray~\citep{hoyer2017xarray} provide powerful abstractions
for large-scale data analysis and can be
used to perform interactive analysis
of large VCF Zarr datasets,
with computations transparently distributed over a cluster.
As shown in our benchmarks, lower-level computations
in which optimised kernels are applied on a chunk-by-chunk
basis are also straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers can 
choose the appropriate level of abstraction and tool chain
on a case-by-case basis. These abstractions will extend, hopefully, to
include more convenient primitives for genomics workflows through
projects like sgkit in the future.
